Asteroseismology: AI-Driven Seismic Event Detection for Extra-Terrestrial Missions\\

### 2025-05-25 paper.tex
update

\section{Background}
    \subsection{Lunar Seismology}
        The foundation of extra-terrestrial seismology was laid by the NASA Apollo missions (1969-1977). The Apollo
        Lunar Surface Experiments Package (ALSEP) successfully deployed a network of four primary seismic stations
        (Apollo 12, 14, 15, and 16), complemented by the short-lived Apollo 11 seismometer and the Lunar Seismic
        Profiling Experiment (LSPE) geophone array on Apollo 17 \cite{Nakamura1982, Lognonne2005, Nunn2020}
        %(Nakamura et al., 1981; Lognonné, 2005; Nunn et al., 2020a).
        These Passive Seismic Experiments (PSE) typically included three-component long-period (LP, later referred to as
        mid-period or MP) seismometers and a single-component short-period (SP) seismometer
        \cite{Nunn2020}%(Nunn et al., 2020a).

@@

        \subsubsection{CNN Architecture}
            The core of the MLEV stage is a lightweight CNN designed for binary classification (event vs. noise). The
            architecture, implemented using Keras \cite{Chollet2021}, is structured to process the seismogram and
            auxiliary inputs efficiently:

@@

@@

        \subsection{Datasets and Input Seismogram Preparation}
            hello
            \subsubsection{Data Sources (Lunar and Martian)}
                hello
                \begin{enumerate}[label=\alph*), leftmargin=3em]
                    \item \textit{Apollo ALSEP Data:} This dataset comprises seismic recordings f  rom the Apollo Lunar
                    Surface Experiments Package, offering crucial insights into lunar seismicity and the Moon's internal
                    structure.
                    \item \textit{Mars InSight Mission Data:} Waveforms from NASA's Mars InSight mission, specifically from
                    the SEIS instrument, provide a wealth of information on current Martian seismic activity and atmospheric
                    phenomena.
                \end{enumerate}
            \subsubsection{Waveform Segmentation and Initial Processing}
                hello

        \section{Related Work}
\label{sec:related_work}
The detection of seismic events from continuous waveform data is a fundamental task in seismology. For decades, this has been addressed using a variety of algorithmic approaches, evolving from simple amplitude thresholding to more sophisticated statistical and pattern recognition techniques.

\subsection{Traditional Seismic Event Detection}
The STA/LTA algorithm, first comprehensively described by Allen \cite{b14} and further developed by others \cite{b26}, remains one of the most widely used methods for automatic seismic event detection due to its simplicity and computational efficiency. It operates by comparing the short-term average of the signal amplitude (or power) to its long-term average, triggering a detection when this ratio exceeds a predefined threshold. While effective for signals with clear onsets and high signal-to-noise ratios (SNR), STA/LTA algorithms are susceptible to false triggers from transient noise, instrument glitches, or complex signals with emergent onsets, particularly in the challenging noise environments encountered on other planetary bodies \cite{b15}.

Template matching, or waveform cross-correlation \cite{b18}, offers higher sensitivity for detecting weak, repetitive seismic events, provided a known template waveform exists. This method is particularly useful for identifying families of similar earthquakes or moonquakes. However, its performance heavily depends on the quality and representativeness of the template library and can be computationally intensive for continuous scanning with a large number of templates.

\subsection{Machine Learning in Seismology}
The advent of machine learning has opened new frontiers in seismic data analysis. Early applications involved traditional ML algorithms such as Support Vector Machines (SVMs) and Random Forests for phase picking and event classification \cite{b19}. However, deep learning models, especially CNNs, have demonstrated superior performance in handling raw seismic waveforms and learning complex features directly from the data \cite{b16}.

CNNs have been successfully applied to earthquake detection, phase picking, and magnitude estimation on Earth \cite{b17, b20}. Their ability to learn hierarchical features from time-series or time-frequency representations (like spectrograms) makes them well-suited for distinguishing subtle seismic signals from background noise.

\subsection{Machine Learning in Planetary Seismology}
The application of ML to planetary seismology is a more recent but rapidly growing field, driven by the unique challenges of these missions. The work by Civilini et al. (2021) \cite{favPaper} is a significant contribution in this area. They developed CNN models trained on spectrograms of seismic events from a single Earth station and successfully applied these, using transfer learning, to detect moonquakes in the Apollo Passive Seismic Experiment (PSE) and Lunar Seismic Profiling Experiment (LSPE) data. Their approach demonstrated the feasibility of using non-local training data and highlighted the potential for CNNs to catalog planetary seismicity even without prior local seismic data. They also introduced an "extra-arrival accuracy metric" to quantify performance on noisy lunar datasets.

@@

\section{The Asteroseismology Framework}

\subsection{Performance Results}
Initial evaluations of the Asteroseismology framework demonstrate promising results. The CNN component, even with what is currently considered limited training data and refinement, achieved a **detection accuracy of over 80\%** on the test set comprising both lunar and Martian seismic signals. This indicates the model's capability to generalize across different planetary environments and event characteristics.

Table \ref{tab:performance_summary} (placeholder) would typically summarize these key performance indicators. Figure \ref{fig:roc_curve} (placeholder) would show the Receiver Operating Characteristic (ROC) curve for the CNN classifier.

\begin{table}[htbp]
    \caption{Preliminary Performance Summary of Asteroseismology (Illustrative)}
    \begin{center}
        \begin{tabular}{|l|c|}
            \hline
            \textbf{Metric}                 & \textbf{Value} \\
            \hline
            Detection Accuracy (Events)     & $>$80\%        \\
            False Positive Rate (Noise)     & $\sim$5\%      \\
            Precision                       & [e.g., 0.85]   \\
            Recall                          & [e.g., 0.82]   \\
            F1-Score                        & [e.g., 0.83]   \\
            Processing Speed (1 month data) & $<$60 seconds  \\
            \hline
            \multicolumn{2}{l}{Note: Values are indicative based on abstract; actual results needed.}
        \end{tabular}
        \label{tab:performance_summary}
    \end{center}
\end{table}

@@

\subsection{Adaptability and Tunability}
The six tunable parameters within the framework allow for significant adaptation to different mission requirements. For example, on a mission expecting very subtle seismic signals, the STA/LTA trigger thresholds and characteristic function parameters can be adjusted for higher sensitivity, potentially at the cost of passing more candidates to the CNN. Conversely, in a very noisy environment or when bandwidth is extremely limited, parameters can be set more conservatively to prioritize only the clearest events. This flexibility is a key advantage for diverse planetary targets and mission phases.

\subsection{Discussion}
The preliminary results suggest that the Asteroseismology framework offers a compelling solution for autonomous seismic event detection in space missions. The hybrid approach appears to strike an effective balance: the conventional algorithms provide a computationally cheap first pass that significantly reduces the data volume and false alarm rate for the CNN, while the CNN provides the sophisticated pattern recognition needed to distinguish true, often subtle, seismic events from complex noise.

Compared to purely conventional methods, Asteroseismology offers substantially lower false positive rates. When compared to ML approaches like that of Civilini et al. \cite{favPaper}, which primarily used spectrograms as CNN input, our use of 1D waveforms combined with auxiliary statistical features offers a different pathway for feature extraction. While a direct quantitative comparison is complex without reimplementing their specific model on our exact data splits, our approach aims to preserve fine temporal details in the waveform and provide explicit, interpretable auxiliary features to the CNN, which could be advantageous for certain event types or computational constraints. The processing of data from both lunar (ALSEP) and Martian (InSight) missions in our training and testing pipeline also demonstrates a broader applicability than studies focused solely on one body.

The current accuracy of $>$80\% with limited training is encouraging. Further expansion of the training dataset, including more diverse event types and noise conditions, along with more sophisticated data augmentation, is expected to further improve this performance. The planned integration with an AC-GAN framework (discussed in Future Work) is also anticipated to boost robustness.

One limitation of the current study is the reliance on existing event catalogs for labeling, which may have their own biases or incompleteness. Future work could involve semi-supervised learning approaches to leverage larger amounts of unlabeled data.

@@

Future work will focus on several key areas to further enhance the Asteroseismology framework:
\begin{enumerate}
    \item \textbf{CNN Architecture and Training Refinement:} We will explore more advanced CNN architectures and expand the training dataset significantly. This will include incorporating a wider variety of seismic event types, noise profiles, and employing sophisticated data augmentation techniques to improve generalization and robustness.
    \item \textbf{Integration with Auxiliary Classifier Conditional GAN (AC-GAN):} A primary future development is the refinement of the CNN within an AC-GAN framework \cite{b22}. AC-GANs can be used to generate more realistic synthetic seismic data for training, particularly for rare event types, and can also improve the classifier's ability to distinguish subtle differences between event classes and noise by simultaneously learning to classify and generate data. This is expected to further improve detection reliability and reduce false positives.
    \item \textbf{Enhanced Auxiliary Features:} We will investigate the inclusion of a richer set of physics-informed auxiliary features for the CNN, derived from more detailed analysis of the candidate event waveforms by the conventional algorithms.
    \item \textbf{Onboard Implementation Prototyping:} Efforts will be made to prototype the system on hardware representative of spacecraft processors (e.g., FPGAs or radiation-hardened CPUs) to rigorously evaluate its performance under realistic resource constraints.
    \item \textbf{Expansion to Other Planetary Bodies:} While currently focused on lunar and Martian data, the framework's adaptable design makes it a candidate for future missions to other seismically active bodies like Europa or Titan, with appropriate tuning and training data.
\end{enumerate}

In conclusion, the Asteroseismology framework represents a significant step towards enabling more autonomous and scientifically productive planetary seismology missions. By intelligently managing data at the source, it helps to overcome critical mission constraints, paving the way for deeper and more comprehensive exploration of the internal structures of celestial bodies in our solar system.

>>>>

\section{Background}
    The detection of seismic events from continuous waveform data is a fundamental task in seismology. For decades, this
    has been addressed using a variety of algorithmic approaches, evolving from simple amplitude thresholding to more
    sophisticated statistical and pattern recognition techniques.
    \subsection{Lunar Seismology}
        The foundation of extra-terrestrial seismology was laid by the NASA Apollo missions (1969-1977). The Apollo
        Lunar Surface Experiments Package (ALSEP) successfully deployed a network of four primary seismic stations
        (Apollo 12, 14, 15, and 16), complemented by the short-lived Apollo 11 seismometer and the Lunar Seismic
        Profiling Experiment (LSPE) geophone array on Apollo 17 \cite{Nakamura1982, Lognonne2005, Nunn2020}
        %(Nakamura et al., 1981; Lognonné, 2005; Nunn et al., 2020a).
        These Passive Seismic Experiments (PSE) typically included three-component long-period (LP, later referred to as
        mid-period or MP) seismometers and a single-component short-period (SP) seismometer
        \cite{Nunn2020}%(Nunn et al., 2020a).

@@

        \subsubsection{CNN Architecture}
            The core of the MLEV stage is a CNN designed for binary classification (event vs. noise). The
            architecture, implemented using Keras \cite{Chollet2021} and visualized in ``Fig.~\ref{fig:cnn_architecture_diagram}'', is structured to process the seismogram and
            auxiliary inputs:

@@

            \begin{figure}[htbp]
                \centerline{\includegraphics[width=0.7\columnwidth]{figures/CNN_Figure.png}}
                \caption{Detailed architecture of the Convolutional Neural Network (CNN) used in the MLEV stage. 
                The model takes two inputs: a 1D seismogram segment (input-3) and auxiliary features (input-4). 
                The seismogram undergoes three convolutional blocks (Conv1D, MaxPooling1D, Dropout) before being 
                flattened. The flattened seismogram features are then concatenated with the auxiliary inputs. 
                This combined representation is processed by two dense layers with dropout, leading to a final sigmoid output layer for binary classification.}
                \label{fig:cnn_architecture_diagram}
            \end{figure}

@@

        \subsection{Datasets and Input Seismogram Preparation}
            \subsubsection{Data Sources (Lunar and Martian)}
                \begin{enumerate}[label=\alph*), leftmargin=3em]
                    \item \textit{Apollo ALSEP Data:} This dataset comprises seismic recordings from the Apollo Lunar
                    Surface Experiments Package, offering crucial insights into lunar seismicity and the Moon's internal
                    structure.
                    \item \textit{Mars InSight Mission Data:} Waveforms from NASA's Mars InSight mission, specifically from
                    the SEIS instrument, provide a wealth of information on current Martian seismic activity and atmospheric
                    phenomena.
                \end{enumerate}
                % \subsubsection{Waveform Segmentation and Initial Processing}
        \subsection{Performance Results}
        Initial evaluations of the framework demonstrate promising results. The CNN component, even with what is
        currently considered limited training data and refinement, achieved a **detection accuracy of over 80\%** on the
        test set comprising both lunar and Martian seismic signals. This indicates the model's capability to generalize
        across different planetary environments and event characteristics.

        % Table \ref{tab:performance_summary} (placeholder) would typically summarize these key performance indicators.
        % Figure \ref{fig:roc_curve} (placeholder) would show the Receiver Operating Characteristic (ROC) curve for the
        % CNN classifier.

        % \begin{table}[htbp]
        %     \caption{Preliminary Performance Summary of Asteroseismology (Illustrative)}
        %     \begin{center}
        %         \begin{tabular}{|l|c|}
        %             \hline
        %             \textbf{Metric}                 & \textbf{Value} \\
        %             \hline
        %             Detection Accuracy (Events)     & $>$80\%        \\
        %             False Positive Rate (Noise)     & $\sim$5\%      \\
        %             Precision                       & [e.g., 0.85]   \\
        %             Recall                          & [e.g., 0.82]   \\
        %             F1-Score                        & [e.g., 0.83]   \\
        %             Processing Speed (1 month data) & $<$60 seconds  \\
        %             \hline
        %             \multicolumn{2}{l}{Note: Values are indicative based on abstract; actual results needed.}
        %         \end{tabular}
        %         \label{tab:performance_summary}
        %     \end{center}
        % \end{table}

% \section{Related Work}
% \label{sec:related_work}

% \subsection{Traditional Seismic Event Detection}
% The STA/LTA algorithm, first comprehensively described by Allen \cite{b14} and further developed by others \cite{b26},
% remains one of the most widely used methods for automatic seismic event detection due to its simplicity and
% computational efficiency. It operates by comparing the short-term average of the signal amplitude (or power) to its
% long-term average, triggering a detection when this ratio exceeds a predefined threshold. While effective for signals
% with clear onsets and high signal-to-noise ratios (SNR), STA/LTA algorithms are susceptible to false triggers from
% transient noise, instrument glitches, or complex signals with emergent onsets, particularly in the challenging noise
% environments encountered on other planetary bodies \cite{b15}.

% Template matching, or waveform cross-correlation \cite{b18}, offers higher sensitivity for detecting weak, repetitive
% seismic events, provided a known template waveform exists. This method is particularly useful for identifying families
% of similar earthquakes or moonquakes. However, its performance heavily depends on the quality and representativeness of
% the template library and can be computationally intensive for continuous scanning with a large number of templates.

% \subsection{Machine Learning in Seismology}
% The advent of machine learning has opened new frontiers in seismic data analysis. Early applications involved
% traditional ML algorithms such as Support Vector Machines (SVMs) and Random Forests for phase picking and event
% classification \cite{b19}. However, deep learning models, especially CNNs, have demonstrated superior performance in
% handling raw seismic waveforms and learning complex features directly from the data \cite{b16}.

% CNNs have been successfully applied to earthquake detection, phase picking, and magnitude estimation on Earth \cite{b17,
% b20}. Their ability to learn hierarchical features from time-series or time-frequency representations (like
% spectrograms) makes them well-suited for distinguishing subtle seismic signals from background noise.

% \subsection{Machine Learning in Planetary Seismology}
% The application of ML to planetary seismology is a more recent but rapidly growing field, driven by the unique
% challenges of these missions. The work by Civilini et al. (2021) \cite{Civilini2021} is a significant contribution in this
% area. They developed CNN models trained on spectrograms of seismic events from a single Earth station and successfully
% applied these, using transfer learning, to detect moonquakes in the Apollo Passive Seismic Experiment (PSE) and Lunar
% Seismic Profiling Experiment (LSPE) data. Their approach demonstrated the feasibility of using non-local training data
% and highlighted the potential for CNNs to catalog planetary seismicity even without prior local seismic data. They also
% introduced an "extra-arrival accuracy metric" to quantify performance on noisy lunar datasets.

@@

@@

% \subsection{Adaptability and Tunability}
% The six tunable parameters within the framework allow for significant adaptation to different mission requirements. For example, on a mission expecting very subtle seismic signals, the STA/LTA trigger thresholds and characteristic function parameters can be adjusted for higher sensitivity, potentially at the cost of passing more candidates to the CNN. Conversely, in a very noisy environment or when bandwidth is extremely limited, parameters can be set more conservatively to prioritize only the clearest events. This flexibility is a key advantage for diverse planetary targets and mission phases.

% \subsection{Discussion}
% The preliminary results suggest that the Asteroseismology framework offers a compelling solution for autonomous seismic
% event detection in space missions. The hybrid approach appears to strike an effective balance: the conventional
% algorithms provide a computationally cheap first pass that significantly reduces the data volume and false alarm rate
% for the CNN, while the CNN provides the sophisticated pattern recognition needed to distinguish true, often subtle,
% seismic events from complex noise.

% Compared to purely conventional methods, Asteroseismology offers substantially lower false positive rates. When compared
% to ML approaches like that of Civilini et al. \cite{favPaper}, which primarily used spectrograms as CNN input, our use
% of 1D waveforms combined with auxiliary statistical features offers a different pathway for feature extraction. While a
% direct quantitative comparison is complex without reimplementing their specific model on our exact data splits, our
% approach aims to preserve fine temporal details in the waveform and provide explicit, interpretable auxiliary features
% to the CNN, which could be advantageous for certain event types or computational constraints. The processing of data
% from both lunar (ALSEP) and Martian (InSight) missions in our training and testing pipeline also demonstrates a broader
% applicability than studies focused solely on one body.

% The current accuracy of $>$80\% with limited training is encouraging. Further expansion of the training dataset,
% including more diverse event types and noise conditions, along with more sophisticated data augmentation, is expected to
% further improve this performance. The planned integration with an AC-GAN framework (discussed in Future Work) is also
% anticipated to boost robustness.

% One limitation of the current study is the reliance on existing event catalogs for labeling, which may have their own
% biases or incompleteness. Future work could involve semi-supervised learning approaches to leverage larger amounts of
% unlabeled data.

@@

Future work can focus on several key areas to further enhance the framework:
\begin{enumerate}
    \item \textbf{CNN Architecture and Training Refinement:} Exploring more advanced CNN architectures and expand
    the training dataset significantly. This will include incorporating a wider variety of seismic event types, noise
    profiles, and employing sophisticated data augmentation techniques to improve generalization and robustness.
    \item \textbf{Integration with Auxiliary Classifier Conditional GAN (AC-GAN):} A primary future development is the
    refinement of the CNN within an AC-GAN framework \cite{b22}. AC-GANs can be used to generate more realistic
    synthetic seismic data for training, particularly for rare event types, and can also improve the classifier's
    ability to distinguish subtle differences between event classes and noise by simultaneously learning to classify and
    generate data. This is expected to further improve detection reliability and reduce false positives.
    \item \textbf{Enhanced Auxiliary Features:} We will investigate the inclusion of a richer set of physics-informed
    auxiliary features for the CNN, derived from more detailed analysis of the candidate event waveforms by the
    conventional algorithms.
    \item \textbf{Onboard Implementation Prototyping:} Efforts will be made to prototype the system on hardware
    representative of spacecraft processors (e.g., FPGAs or radiation-hardened CPUs) to rigorously evaluate its
    performance under realistic resource constraints.
    \item \textbf{Expansion to Other Planetary Bodies:} While currently focused on lunar and Martian data, the
    framework's adaptable design makes it a candidate for future missions to other seismically active bodies like Europa
    or Titan, with appropriate tuning and training data.
\end{enumerate}

In conclusion, the framework represents a significant step towards enabling more autonomous and scientifically
productive planetary seismology missions. By intelligently managing data at the source, it helps to overcome critical
mission constraints, paving the way for deeper and more comprehensive exploration of the internal structures of
celestial bodies in our solar system.
